Processing genome scale tabular data with wormtable

### 2013-07-29 docs/paper/paper.tex
sorted out sectioning a little.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Implementation 
%%
\section*{Implementation}
\emph{This should include a description of the overall architecture of the software implementation, 
along with details of any critical issues and how they were addressed.
}
% what is wormtable?
Wormtable (write-once read-many table) 
is a new data format and software library designed specifically
to deal with the challenges of data processing in bioinformatics.
It provides a portable, compact, read-only store for tabular
data of essentially unlimited size. Tables can
be either written directly via the library, or converted
from existing formats via command line tools. 
Wormtable provides a simple Python API to access rows
in the table, facilitating efficient and convenient access to 
data.

% Wormtable's binary row format 
Data is stored using a  
compact binary format designed specifically for biological applications.
Each row consists of a fixed number of columns, and each column 
holds either a fixed or variable number of elements of a fixed type and size.
Wormtable currently supports signed and unsigned integers, 
floating point and character data. To ensure that values are 
stored as compactly as possible, integer sizes can range from  
$1$ to $8$ bytes. The standard single and double precision 
floating point values are also supported, along with fixed and 
variable length strings.

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Results and Discussion %%
%%
\section*{Results and Discussion}

@@

\subsection*{Examples}
The Variant Call Format (VCF) encodes information about variant sites in a genome 
as tab-delimited rows in a text file~\cite{da+11}.  
To illustrate the advantages of wormtable, we
converted a large 
publicly available VCF file, produced as part of the
Drosophila genetic reference panel~\cite{mr+12}, to wormtable format.
This VCF\cite{freeze.vcf}
%\footnote{\href%
%{ftp://ftp.hgsc.bcm.edu/DGRP/freeze2\_Feb\_2013/vcf\_files/freeze2.vcf.gz}
%{ftp://ftp.hgsc.bcm.edu/DGRP/freeze2\_Feb\_2013/vcf\_files/freeze2.vcf.gz}}
consists of 15GB of uncompressed text and contains data from the 
whole genome over 6,146,611 rows.
Using the included \texttt{vcf2wt} program
the VCF was converted to wormtable format 
on a standard workstation with a single hard disk. The
conversion required approximately 68 minutes, and the size of the resulting  
wormtable file was 10.1GB (using the smallest type required
to represent the data in each column).

@@

@@

It is well known that parsing text files is a very inefficient process, 
and binary file formats are preferable for this reason. The BCF format,
for example, is a binary version of the VCF format discussed above which 
is designed to avoid just this problem. There are significant difficulties,
however, with having many different binary file formats for bioinformatics 
data. Each binary format requires a library and set of tools to view 
and process it, as it is not reasonable to expect users to decode binary
files. Bindings for several different languages must also be provided, if 
the file format is to be widely used.
Maintaining these libraries, tools and language bindings across 
different processor architectures and operating systems
is a complex software engineering task. Maintaining this ecosystem 
separately for several different file formats is surely 
unsustainable.

% If we get some help, it could be pretty awesome...
Wormtable is an open and collaborative project seeking feedback and 
contributors. The project is in its early stages, currently 
providing a conversion tool for the VCF format only. With community
participation it may grow to support many other formats and become
a powerful common platform for bioinformatics data analysis.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Implementation 
%%
\section*{Implementation}
\emph{This should include a description of the overall architecture of the software implementation, 
along with details of any critical issues and how they were addressed.
}

% what is wormtable?
Wormtable (write-once read-many table) 
is a new data format and software library designed specifically
to deal with the challenges of data processing in bioinformatics.
It provides a portable, compact, read-only store for tabular
data of essentially unlimited size. Tables can
be either written directly via the library, or converted
from existing formats via command line tools. 
Wormtable provides a simple Python API to access rows
in the table, facilitating efficient and convenient access to 
data.

% Wormtable's binary row format 
Data is stored using a  
compact binary format designed specifically for biological applications.
Each row consists of a fixed number of columns, and each column 
holds either a fixed or variable number of elements of a fixed type and size.
Wormtable currently supports signed and unsigned integers, 
floating point and character data. To allow data to be stored as 
efficiently as possible, wormtable supports a wide range of 
type sizes: integers range from one to eight bytes long,
and IEEE half, single and double precision floating point 
are supported. Fixed and variable length character strings 
are also supported.

% Tables

@@

% Indexes

@@

% Layout & implementation.
Each table corresponds to a single directory in the file system.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Results and Discussion %%
%%
\section*{Results}

@@

\subsection{API}

\subsection*{Examples}
The Variant Call Format (VCF) encodes information about variant sites in a genome 
as tab-delimited rows in a text file~\cite{da+11}.  
To illustrate the advantages of wormtable, we
converted a large 
publicly available VCF file, produced as part of the
Drosophila genetic reference panel~\cite{mr+12}, to wormtable format.
This VCF\cite{freeze.vcf}
consists of 15GB of uncompressed text and contains data from the 
whole genome over 6,146,611 rows.
Using the included \texttt{vcf2wt} program
the VCF was converted to wormtable format 
on a standard workstation with a single hard disk. The
conversion required approximately 68 minutes, and the size of the resulting  
wormtable file was 10.1GB (using the smallest type required
to represent the data in each column).

@@

\section{Discussion}
It is well known that parsing text files is a very inefficient process, 
and binary file formats are preferable for this reason. The BCF format,
for example, is a binary version of the VCF format discussed above which 
is designed to avoid just this problem. There are significant difficulties,
however, with having many different binary file formats for bioinformatics 
data. Each binary format requires a library and set of tools to view 
and process it, as it is not reasonable to expect users to decode binary
files. Bindings for several different languages must also be provided, if 
the file format is to be widely used.
Maintaining these libraries, tools and language bindings across 
different processor architectures and operating systems
is a complex software engineering task. Maintaining this ecosystem 
separately for several different file formats is surely 
unsustainable.

The library supports efficient access to any data stored in 
wormtable format, but there is currently a conversion tool for
the VCF format only. The most important aspect of future development 
is to develop tools to convert other tabular formats such as 
PSL, GFF, and SAM to wormtable format [Dan, Rob: does this make sense?
are there other formats we should chuck in here?]. Since the conversion 
tool can be written entirely in Python this is not a difficult task.

Other probable directions for future development of wormtable would
be to incorporate BGZF compression~\cite{li11}, allowing for smaller
files with efficient random access; to develop a 
C library and bindings for other popular languages such as 
Perl and R; and to generally enrich the API with features useful 
to the community. 
Wormtable is an open and collaborative project, which would welcome 
feedback and contributions from the community

@@
